Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-21 sections/eval.tex
Update on Overleaf.

\subsubsection{SpMV}
\willow{We might want to start this section off by saying spmv is a bandwidth bound kernel, and historically therefore subject to many format modifications to improve throughput and reduce the associated control flow}
Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data. More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats. \willow{say this faster} Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.

Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns. \willow{this might show up in the intro, but if you wanna say this, maybe give an exmaple here too (e.g. symmetry)} Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.

\willow{these paragraphs could be merged and condensed. Perhaps some of this belongs elsewhere} The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite]. \willow{The following two sentences are the most important and relevant so far} The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats. In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.

\willow{fix the citations, put Figure \ref{fig:spmv} earlier, We display speedup relative to ... } Figures \ref{spmv_sorted} and \ref{spmv_grouped} compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library. We display speedup relative to row major or column major TACO performance, whichever was faster. We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface [cite], Ahrens et al. to test a variable block row format partitioning strategy [cite], and Kjolstad et al. to test the TACO library [cite]. In addition, we also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases. The dense vector is randomly generated. We tested a wide variety of finch formats, described in the following subsection; the performance displayed for Finch on each dataset is the fastest among the formats we tested.

@@

\subsubsection{Tensor Formats}
For the tensor formats in Table \ref{spmv_tensor_formats}, we tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch. Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.

\begin{table}[htbp]
    \centering
    \label{spmv_tensor_formats}
    \caption{SpMV Tensor Formats}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
        \hline
        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
        \cline{3-4} 
        & & Pattern & sparse, boolean-valued matrices \\
        \cline{2-4} 
        & SparseVBL & Element & real-valued matrices with blocked structure \\
        \cline{2-4}
        & SparseBand & Element & real-valued matrices with diagonal band \\
        \cline{2-4}
        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
        \cline{3-4} 
        & & Pattern & matrix with runs of true or false \\
        \hline 
    \end{tabular}

@@

\subsubsection{4D Blocked SpMV}
Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks. Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size. We represent the blocked matrix with a SparseList as its second level (i.e. of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored. Then, we perform SpMV on each b x b block individually. Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block. This method enables us to take advantage of spatial and temporal locality via register reuse [cite].

We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector. We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.

Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications. In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite]. Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected. For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].

@@

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
    \label{spmv_grouped}
\end{figure}

>>>>

\subsubsection{SpMV}
The sparse matrix-vector multiplication kernel (SpMV) is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite]. SpMV is a bandwidth bound kernel, and thus historically has been subject to many modifications to improve throughput and reduce the associated control flow [cite]. The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.

In this case study, we highlight a few different Finch formats, as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do. Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).

We display speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library, as depicted in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface [cite], Ahrens et al. to test a variable block row format partitioning strategy [cite], and Kjolstad et al. to test the TACO library [cite]. Additionally, we included the SNAP graph collection to test with boolean matrices. We also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases. The dense vector is randomly generated. We tested a wide variety of Finch formats; the performance displayed for Finch on each dataset is the fastest among the formats we tested.

@@

\subsubsection{Tensor Formats}
For the tensor formats in Table \ref{spmv_tensor_formats}, we tested with both the row-major and the column-major implementations with Finch. Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.

\begin{table}[htbp]
    \centering
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
        \hline
        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
        \cline{3-4} 
        & & Pattern & sparse, boolean-valued matrices \\
        \cline{2-4} 
        & SparseVBL & Element & real-valued matrices with blocked structure \\
        \cline{2-4}
        & SparseBand & Element & real-valued matrices with diagonal band \\
        \cline{2-4}
        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
        \cline{3-4} 
        & & Pattern & matrix with runs of true or false \\
        \hline 
    \end{tabular}

@@

% Perhaps add the following section again in a future iteration of paper
% \subsubsection{4D Blocked SpMV}
% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks. Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size. We represent the blocked matrix with a SparseList as its second level (i.e. of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored. Then, we perform SpMV on each b x b block individually. Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block. This method enables us to take advantage of spatial and temporal locality via register reuse [cite].

% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector. We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.

% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications. In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite]. Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected. For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].

@@

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
    \label{spmv_grouped}
    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested. "R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
\end{figure}

### 2024-03-21 sections/lang.tex
Update on Overleaf.

\[
    \frac{}{\langle\finchliteral(val), (e, t)\rangle \rightarrow val} \quad
    \frac{}{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle} \quad
\]
\[
    \frac{}{\langle\finchvar(name), (e, t)\rangle \rightarrow e(\finchvar(name))} \quad
    \frac{}{\langle\finchindex(name), (e, t)\rangle \rightarrow e(\finchindex(name))}
\]
\[
    \frac{\langle args_i, (e, t) \rangle \Rightarrow vals_i \quad \langle f, (e, t) \rangle \Rightarrow g}
    {\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
\]
\[
\frac{\langle head, s_1 \rangle \rightarrow s_2}
{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle} \quad
\frac{}{\langle\finchblock(), s\rangle \rightarrow s}
\]
\[
\frac{\langle cond, s \rangle \Rightarrow true}
{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}\quad
\frac{\langle cond, s \rangle \Rightarrow false}
{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
\]
\[
\frac{\langle val, (e, t) \rangle \rightarrow val'}
{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
\]
\[
\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns' \quad  \langle init, s \rangle \rightarrow init' \quad  \langle dims_i, s \rangle \rightarrow dims_i'}
{\langle\finchdeclare(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle declare(tns', init', dims'...), t \rangle)}
\]
\[
\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns'}
{\langle\finchthaw(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
\]
\[
\frac{s = (e, t) \quad e(\finchmode(tns)) = update \quad e(tns) = tns'}
{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
\]

@@

\[
\frac{e(tns) = tns' \langle unfurl(tns'), t\rangle \rightarrow \langle tns'', t'\rangle}
{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), (e, t)\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), (e, t')\rangle}
\]

>>>>

\begin{figure}
    \centering
    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = read}
    \hypo{e(tns) = tns'}
    \infer3{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
    \end{prooftree}
    \hfill
    \vspace{2em}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \infer3{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
    \end{prooftree}
      \vspace{2em}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{tns\not\in e}
    \hypo{\langle init, s \rangle \Rightarrow v}
    \hypo{\forall d_{i}\in dims, \langle init, d_{i} \rangle \Rightarrow v_{i}}
    \infer4{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns, v, v_{0}, \dots)), t \rangle)}
    \end{prooftree}

    \caption{Rules for evaluating life cycle functions.}
    \label{fig:enter-label}
\end{figure}

\begin{figure}
    \centering
    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \infer1{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
    \end{prooftree}
    \vspace{2em}
    \caption{Basic Evaluation Rules}
    \label{fig:enter-label}
\end{figure}

\[
    \frac{}{\langle\finchliteral(val), (e, t)\rangle \rightarrow val} \quad
    \frac{}{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle} \quad
\]
\[
    \frac{}{\langle\finchvar(name), (e, t)\rangle \rightarrow e(\finchvar(name))} \quad
    \frac{}{\langle\finchindex(name), (e, t)\rangle \rightarrow e(\finchindex(name))}
\]
\[
    \frac{\langle args_i, (e, t) \rangle \Rightarrow vals_i \quad \langle f, (e, t) \rangle \Rightarrow g}
    {\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
\]
\[
\frac{\langle head, s_1 \rangle \rightarrow s_2}
{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle} \quad
\frac{}{\langle\finchblock(), s\rangle \rightarrow s}
\]
\[
\frac{\langle cond, s \rangle \Rightarrow true}
{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}\quad
\frac{\langle cond, s \rangle \Rightarrow false}
{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
\]

@@
